Sparse Function-space Representation of Neural Networks

### 2023-09-04 workshops/icml-duality/main.tex
intro stuff

\section{Introduction}
%
Deep learning \citep{goodfellow2016deep} has become the cornerstone of contemporary artificial intelligence, proving remarkably effective in tackling supervised and unsupervised learning tasks in the {\em large data}, {\em offline}, and {\em gradient-based training} regime. Despite its success, gradient-based learning techniques exhibit limitations. Firstly, how can we efficiently quantify uncertainty without resorting to expensive and hard-to-interpret sampling in the model's weight-space? Secondly, how to update the weights of an already trained model with new batches of data without compromising the performance on past data? These questions become central when applied to sequential learning paradigms, such as continual learning \citep[CL,][]{parisi2019continual, de2021continual} and reinforcement learning  \citep[RL,][]{sutton2018reinforcement}. In CL, access to the previous data is lost, and then the challenge is retaining a compact representation of the problem to alleviate forgetting over the life-long learning horizon~\citep{mccloskey1989catastrophic}.
Similarly, in RL, the model must adapt to environmental observations and can leverage predictive uncertainties for exploration.\looseness-1

@@

%Current state of affairs
Recent techniques  \citep[\eg,][]{ritter2018kfac,khan2019approximate,daxberger2021laplace,fortuin2021bayesian,immer2021scalable} apply a Laplace-GGN approximation to convert trained NNs into Bayesian neural networks (BNNs), that can provide uncertainty without sacrificing additional resources to training \citep{foong2019between}. Furthermore, the resultant weight-space posterior can be converted to the function-space as shown in \citet{khan2019approximate, immer2021improving}. The function-space representation allows for a principled mathematical approach for analyzing its behaviour \citep{cho2009kernel,meronen2020stationary}, performing probabilistic inference \citep{khan2019approximate}, and quantifying uncertainty in NNs \citep{foong2019between}. These methods rely on the linearization of the NN and the resultant neural tangent kernel \citep[NTK,][]{jacot2018neural}. The NN is characterized in function-space by its first two moment functions, a mean function and covariance function (or kernel)---defining a Gaussian process \citep[GP,][]{rasmussen2006gaussian}. GPs provide a widely used probabilistic toolkit with principled uncertainty estimates. They serve as a standard surrogate model for Bayesian optimization \citep{garnett_bayesoptbook_2022} and are effective in model-based reinforcement learning \citep{deisenroth2011pilco} with theoretical guarantees on regret bounds \citep{srinivas2009gaussian}.
%Yet many problems lie in high dimensional input space; for example, images are where GPs cannot learn representations. In such scenarios such as in many reinforcement learning environments neural networks are used as the surrogate model. However, uncertainty is still essential to ensure effective exploration for sequential algorithms. Successful approaches have attempted to blend neural networks with uncertainty estimates around predictions, allowing for sophisticated exploration strategies. However, there has been limited use of hybrid models that possess the feature representation ability of neural networks but also attractive the properties of GPs, such a hybrid method we propose in this paper.

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that the NN emits `dual' parameters which are the building blocks of a GP~\citep{csato2002sparse, adam2021dual, chang2023memory} . In contrast to previous work that utilizes subsets of training data \citep{immer2021scalable}, this parameterization allows capturing the contributions from {\em all} data points into a sparse representation, essential for predictive uncertainty. Crucially, the resulting GP directly predicts in the same space as the original trained NN, with the benefit of avoiding the complexity introduced by working in weight-space and the notorious cubic complexity of vanilla GPs.
%\todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}
%, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs.
Through the dual parameter formulation, we establish a connection between the NN, full GPs, and a sparse approximation similar to sparse variational GPs~\citep{titsias2009variational}. We refer to our method as Sparse Function-space Representation (\our)---a sparse GP derived from a trained NN; see \cref{fig:teaser} for an example. %Moreover, this dual parameterization can be exploited to perform dual conditioning \citep{chang2022fantasizing}, \ie, an effective approach for conditioning on new data without needing to retrain the model (see \cref{fig:teaser}).
%As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.
\looseness-2

>>>>

\section{Introduction}
%
Deep learning \citep{goodfellow2016deep} has become the cornerstone of contemporary artificial intelligence, proving remarkably effective in tackling supervised and unsupervised learning tasks in the {\em large data}, {\em offline}, and {\em gradient-based training} regime. Despite its success, gradient-based learning techniques exhibit limitations. Firstly, how can we efficiently quantify uncertainty without resorting to expensive and hard-to-interpret sampling in the model's weight-space? Secondly, how to update the weights of an already trained model with new batches of data without compromising the performance on past data?
These questions become central when applied to sequential learning paradigms, such as continual learning \citep[CL,][]{parisi2019continual, de2021continual}, Bayesian optimization \citep[BO,][]{garnett_bayesoptbook_2022} and reinforcement learning  \citep[RL,][]{sutton2018reinforcement}.
% In CL, access to the previous data is lost, and then the challenge is retaining a compact representation of the problem to alleviate forgetting over the life-long learning horizon~\citep{mccloskey1989catastrophic}.
% Similarly, in RL, the model must adapt to environmental observations and can leverage predictive uncertainties for exploration.\looseness-1

@@

%Current state of affairs
Recent techniques  \citep[\eg,][]{ritter2018kfac,khan2019approximate,daxberger2021laplace,fortuin2021bayesian,immer2021scalable} apply a Laplace-GGN approximation to convert trained NNs into Bayesian neural networks (BNNs), that can provide uncertainty without sacrificing additional resources to training \citep{foong2019between}. Furthermore, the resultant weight-space posterior can be converted to the function-space as shown in \citet{khan2019approximate, immer2021improving}. The function-space representation allows for a principled mathematical approach for analyzing its behaviour \citep{cho2009kernel,meronen2020stationary}, performing probabilistic inference \citep{khan2019approximate}, and quantifying uncertainty in NNs \citep{foong2019between}. These methods rely on the linearization of the NN and the resultant neural tangent kernel \citep[NTK,][]{jacot2018neural}.
The NN is characterized in function-space by its first two moment functions, a mean function and covariance function (or kernel)---defining a Gaussian process \citep[GP,][]{rasmussen2006gaussian}. GPs provide a widely used probabilistic toolkit with principled uncertainty estimates.
% They serve as a standard surrogate model for Bayesian optimization \citep{garnett_bayesoptbook_2022} and are effective in model-based reinforcement learning \citep{deisenroth2011pilco} with theoretical guarantees on regret bounds \citep{srinivas2009gaussian}.
%Yet many problems lie in high dimensional input space; for example, images are where GPs cannot learn representations. In such scenarios such as in many reinforcement learning environments neural networks are used as the surrogate model. However, uncertainty is still essential to ensure effective exploration for sequential algorithms. Successful approaches have attempted to blend neural networks with uncertainty estimates around predictions, allowing for sophisticated exploration strategies. However, there has been limited use of hybrid models that possess the feature representation ability of neural networks but also attractive the properties of GPs, such a hybrid method we propose in this paper.

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that NN's emit `dual' parameters which are the building blocks of a GP~\citep{csato2002sparse, adam2021dual, chang2023memory}.
In contrast to previous work that utilizes subsets of training data \citep{immer2021scalable}, our parameterization captures
information from {\em all} data points in a sparse representation, essential for scaling to deep learning data sets.
% predictive uncertainty.
Importantly, the dual parameterization can be used to 1) sparsify the GP without requiring further optimization
(e.g. variational inference), and 2) can incorporate new data without retraining (in a computationally efficient manner
by conditioning on new data).
% In contrast to previous work that utilizes subsets of training data \citep{immer2021scalable}, our parameterization captures information from {\em all} data points in a sparse representation, essential for predictive uncertainty.
% Importantly, our method predicts in the same space as the original trained NN, with the benefit of avoiding the
% complexity introduced by working in weight-space and the notorious cubic complexity of vanilla GPs.
%\todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}
%, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs.
Through the dual parameterization, we establish a connection between the NN, GPs, and a sparse GP approximations similar to sparse variational GPs~\citep{titsias2009variational}. We refer to our method as Sparse Function-space Representation (\our)---a sparse GP derived from a trained NN; see \cref{fig:teaser} for an example. %Moreover, this dual parameterization can be exploited to perform dual conditioning \citep{chang2022fantasizing}, \ie, an effective approach for conditioning on new data without needing to retrain the model (see \cref{fig:teaser}).
%As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.
\looseness-2
